MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-05-01 rebuttal.txt

Q: In Figure1, the High-End NVIDIA TESLA is either slower or as fast as the consumer grade laptop GPU. Is this purely due to the compilation time (as Figure 2 would suggest, that it makes a difference)? 
A: You are right. You can see for yourself by comparing gpu_out_real.csv and gpu_out_real_a100.csv files in the artifact. We are unsure why the compilation takes longer for high-end GPU than for laptop GPU, after all, it is a proprietary piece of software. Perhaps the loading of shared libraries takes a different amount of time for the operating systems... We added a note in the Performance of MaBoSS.GPU section.

Q: There is no demonstration of how a user would actually deal with the code - especially since the GPU version is based on a hard-coded implementation of the rules. A demonstrative example would help to get an idea, how a user would actually get started, and how deep he needs to dig into the code - especially for a biological audience, accessability of such tools is an important point
A: Respectfully, it is not true that the GPU version is based on any hard-coded implementation. The compilation of the rules happens during the runtime without any user intervention. We included this note in the second paragraph of the Simulation section.

Q: It is stated that synthetic models with up to 1000 nodes were created. The synthetic models themselves are available, however, the description of the chosen models is also important to be mentioned in the paper: I would suggest a table that lists the most important features of the different models (size, nodes, formula sizes, how many different models were actually taken into account, ..) that were created (this table can also include the characteristics of the 3 real-world models to put in perspective). 
A: We included the proposed table in Results section and extended the first paragraph in the Benchmarking Methodology to describe the table fields.

Q: The statistics are described, but how would a user interpret them/make use of them for themselves? What is this shared associative structure that the model is updating? Maybe a diagram would be helpful for visualization?
A: We added a new section Statistics output and visualization, which includes how the statistics output is represented in the code and how a user can interpret and visualize it.

Q: How does the code actually deal with circular steady states?
A: If a model contains circular steady states, simulated trajectories that would indefinitely transition over circular states will eventually reach the maximum time and their simulation stops. Their statistics are further computed without any change. The original paper (Stoll et al.) discusses the ways how to detect circular steady states from the computed statistics, but we think that describing these methods is out of the scope of this paper. Therefore, we included general information about circular steady states in Statistics output and visualization section and referenced the original paper for the reader who wants to know more detailed information.

@@

Q: Figure 3: The x-axis states “Nodes count (log−scale)”, however, it appears to have a linear scale. The linear relationships claimed in the statement “We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).” is not clear from the provided figures, and should be clarified or, for example provide Figure 3 with a linear-linear or log-log scale.
A: Thanks for pointing out the wrong scale in the plot, it is a linear scale. Regarding the linear dependency, it can not be seen from Fig 3 since it shows compilation time percentages rather than the absolute values. Therefore, plotting a linear-linear scale would not show any linear dependency. Nevertheless, the stated linear dependency is indeed there and we observed it from the plots which we did not include in the paper. The general idea why it holds is that increasing the number of nodes by a value of x adds O(x) new lines to a source file which needs to be compiled. We included this explanation in the paper.

Q: I was able to successfully compile MaBoSS with MPI enabled, however I was not able to compile MaBoSS.GPU in a Linux environment.
A: Thanks, compiling with the newest CUDA version revealed a bug in the Thrust dependency. We committed a workaround.

Q: The MaBoSS.GPU README could do with a link to documentation.
A: We added a new section to README with a high-level architectural overview.

Q: Since the authors have disabled issue tracking on their github repositories, I urge them to add a contact method to sysbio-curie/hpcmaboss-artifact and sysbio-curie/MaBoSS.GPU, for users to obtain assistance and provide feedback, including bug reports.
A: This was a mistake, we now have enabled issue tracking on MaBoSS.GPU and hpcmaboss-artifact repositories. Thanks for reminding us.

Q:  It is not clear whether the version provided via the colomoto conda repository is MPI enabled, so a clarification is needed.
A: This is something we have been considering, but we have not moved forward yet. We will most likely have to create specific conda packages which are MPI enabled, in addition to the classic ones. Same thing for MaBoSS.GPU. We also need to think about this for the colomoto docker, but again it might not be straightforward. For now, we think that the target users of these new implementations are knowledgeable enough to install them manually.

Q: A concern is that the provided background is insufficient for the applicability and usefulness of the software to be apparent to the broader audience of BMC Bioinformatics. The authors should add more to provide context and insight into the interpretation and use of MaBoSS output, for those not already familiar with the previous MaBoSS versions.
A: We included a new section Statistics output and visualization.

### 2024-05-01 Article.tex
maboss output section

MaBoSS algorithm simulates the above process to produce stochastic \emph{trajectories}: sequences of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state, and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$. The simulation ends either by a timeout when reaching the maximal allowed time, or by reaching a fixed point state with no outgoing transitions. The algorithm for a single iteration of the trajectory simulation is given explicitly in Algorithm~\ref{alg:iter}.

@@

As an interesting detail, the hash structures pose a surprising constant-factor overhead. In networks where most nodes are internal, the hash map may be replaced by a fixed-size multidimensional array that holds an element for all possible combinations of external node values (basically forming a multidimensional histogram). We discuss the impact of this optimization in the next section.

\subsubsection{MaBoSS CPU Implementation}
MaBoSS was initially developed as a single core application, but very quickly got a basic parallelism to exploit the multi-core nature of modern CPUs. In this parallel implementation the simulation of trajectories, as well as the statistics aggregation, was distributed on multiple cores (controlled by a simulation parameter) using POSIX threads. In the following sections of the papers, we will refer to this implementation as the CPU version.

>>>>

MaBoSS algorithm simulates the above process to produce stochastic \emph{trajectories}: sequences of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state, and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$. The simulation ends either by a timeout when reaching the maximal allowed time, or by reaching a \emph{fixed point} state with no outgoing transitions. The algorithm for a single iteration of the trajectory simulation is given explicitly in Algorithm~\ref{alg:iter}.

@@

As an interesting detail, the hash structures pose a surprising constant-factor overhead. In networks where most nodes are internal, the hash map may be replaced by a fixed-size multidimensional array that holds an element for all possible combinations of external node values (basically forming a multidimensional histogram). We discuss the impact of this optimization in \emph{Implementation} section.

\subsection{MaBoSS CPU Implementation}

MaBoSS was initially developed as a single-core application, but very quickly it was extended with a basic parallelism to exploit the multi-core nature of modern CPUs. In this parallel implementation the simulation of trajectories, as well as the statistics aggregation, was distributed on multiple cores (controlled by a simulation parameter) using POSIX threads. In the following sections of the papers, we will refer to this implementation as the CPU version.

\subsubsection{Statistics output and visualization}

Each of the three kinds of statistics is in its nature a sample from a probabilistic distribution of Boolean states. This sample is represented in code as a hash map in the CPU version, varying in the $(key, value)$ pairs according to the specific statistic. For the final states, the keys of the hash map are the states of the model, and the values are the number of times the state was sampled last in a trajectory. Such output can be visualized as a pie chart (see Figure~\ref{fig:cohen_final}). The fixed states are represented similarly, but only the fixed points are stored in the hash map as keys.

\begin{figure}%[tbhp]
    \centering
    \includegraphics[width=3.25in]{img/cohen_piechart.png}
    \caption{The final states pie chart shows the distribution of the last trajectory states. Labels denote which active non-internal nodes compose the state. \emph{nil} label represents the state where all non-internal nodes are inactive.}
    \label{fig:cohen_final}
\end{figure}

The final and fixed state statistics characterize the behavior of the model at one point in time --- at the end of the simulation. The network state probabilities on a time window highlight more dynamic characteristics of the model, showing how the average trajectory evolves over the simulation time. Programmatically, it is an extension of the final state statistics but instead of one hash map, there is a hash map for each time window. The hash map values are the state durations in the specific time window aggregated over all simulated trajectories. Further, these statistics can be visualized in various ways using a line chart. Figure~\ref{fig:cohen_traj} shows which non-internal nodes are active throughout the simulation.

As mentioned at the beginning of the section, if the trajectory does not reach a fixed point, the simulation is stopped after the maximal allowed time. This is a common scenario especially when some trajectories form circles, i.e. when a model has \emph{circular steady states}. A circular steady state is usually not directly visible from the state probability line charts; Stoll et al.~\cite{stoll2012continuous} proposed methods to detect them (such as plotting the state and transition entropies) but we do not discuss them further in this paper for the sake of brevity.

\begin{figure}
    \centering
    \includegraphics[width=3.25in]{img/cohen_nodes_trajs.png}
    \caption{The line chart of trajectory state probabilities over time windows. Each line represents the ratio of an active non-internal node in the time window over all trajectories (e.g., at the beginning of the simulation, the \emph{Apoptosis} node is inactive in all simulated trajectories and as the time reaches the value of $10$, \emph{Apoptosis} is active in around $40\%$ of trajectories). The x-axis represents the discrete simulation time with the window width of $0.1$.}
    \label{fig:cohen_traj}
\end{figure}
